---
title: "CTA-ED Exercise 3: Comparison and complexity"
author: "Marion Lieutaud"
date: "2/28/2024"
output: html_document
---

## Introduction

The hands-on exercise for this week focuses on: 1) comparing texts; 2) measuring he document-level characteristics of text---here, complexity.

In this tutorial, you will learn how to:
  
* Compare texts using character-based measures of similarity and distance
* Compare texts using term-based measures of similarity and distance
* Calculate the complexity of texts
* Replicate analyses from @schoonvelde_liberals_2019

## Setup 

Before proceeding, we'll load the remaining packages we will need for this tutorial.

```{r, echo=F}
library(kableExtra)
```

```{r, message=F}
library(readr) # more informative and easy way to import data
library(quanteda) # includes functions to implement Lexicoder
library(quanteda.textstats) # for estimating similarity and complexity measures
library(stringdist) # for basic character-based distance measures
library(dplyr) #for wrangling data
library(tibble) #for wrangling data
library(ggplot2) #for visualization

library(tidyverse) # I add this as I need it for the exercise
```


For this example we'll be using data from the 2017-2018 Theresa May Cabinet in the UK. The data are tweets by members of this cabinet. 

```{r}
tweets  <- readRDS(gzcon(url("https://github.com/cjbarrie/CTA-ED/blob/main/data/comparison-complexity/cabinet_tweets.rds?raw=true")))
```

And we see that the data contain three variables: "username," which is the username of the MP in question; "tweet," which is the text of the given tweet, and "date" in days in yyyy-mm-dd format. 

```{r}
head(tweets)
```

And there are 24 MPs whose tweets we're examining. 

```{r}
unique(tweets$username)

length(unique(tweets$username))
```

## Generate document feature matrix

In order to use the `quanteda` package and its accompanying `quanteda.textstats` package, we need to reformat the data into a quanteda "corpus" object. To do this we just need to specify the text we're interested in as well as any associated document-level variables in which we're interested. 

We can do this as follows. 

```{r}
#make corpus object, specifying tweet as text field
tweets_corpus <- corpus(tweets, text_field = "tweet")

#add in username document-level information
docvars(tweets_corpus, "username") <- tweets$username

tweets_corpus
```

We are now ready to reformat the data into a document feature matrix.

```{r}
dfmat <- dfm(tokens(tweets_corpus),
             remove_punct = TRUE, 
             remove = stopwords("english"))

dfmat
```

Note that when we do this we need to have tokenized our corpus object first. We can do this by wrapping the `tokens` function inside the `dfm()` function as above. 

So what is this object? Well the documents here are tweets. And the matrix is a sparse (i.e., mostly zeroes) matrix of 1s and 0s for whether a given word appears in the document (tweet) in question. 

The vertical elements (columns) of this vector are made up of all the words used in all of the tweets combined. Here, it helps to imagine every tweet positioned side by side to understand what's going on here. 

## Compare between MPs

Once we have our data in this format, we are ready to compare between the text produced by members of Theresa May's Cabinet.

Here's an example of the correlations between the combined tweets of 5 of the MPs with each other.

```{r}
corrmat <- dfmat %>%
  dfm_group(groups = username) %>%
  textstat_simil(margin = "documents", method = "correlation")

corrmat[1:5,1:5]

```

Note that here we're using the `dfm_group()` function, which allows you to take a document feature matrix and make calculations while grouping by one of the document-level variables we specified above. 

There are many different measures of similarity, however, that we might think about using.

In the below, we combine four different measures of similarity, and see how they compare to each other across MPs. Note that here we're looking only at the similarity between an MP's tweets and those of then Prime Minister, Theresa May.

## Compare between measures

Let's see what this looks like for one of these measures---cosine similarity.

We first get similarities between the text of MP tweets and all other MPs.

```{r}

# estimate similarity, grouping by username

cos_sim <- dfmat %>%
  dfm_group(groups = username) %>%
  textstat_simil(margin = "documents", method = "cosine") #specify method here as character object

```

But remember we're only interested in how they compare to what Theresa May has been saying. 

So we need to take these cosine similarities and retain only those similarity measures corresponding to the text of Theresa May's tweets. 

We first convert the `textstat_simil()` output to a matrix.

```{r}

cosmat <- as.matrix(cos_sim) #convert to a matrix
  
```

And we can see that the 23rd row of this matrix contains the similarity measures with the Theresa May tweets. 

We take this row, removing the similarity of Theresa May with herself (which will always = 1), and convert it to a datframe object. 

```{r}
#generate data frame keeping only the row for Theresa May
cosmatdf <- as.data.frame(cosmat[23, c(1:22, 24)])
```

We then rename the cosine similarity column with an appropriate name and convert row names to a column variable so that we have cells containing information on the MP to which the cosine similarity measure refers. 

```{r}
#rename column
colnames(cosmatdf) <- "corr_may"
  
#create column variable from rownames
cosmatdf <- tibble::rownames_to_column(cosmatdf, "username")
```

And like so we have our data in tidy format, which we can then plot like so. 

```{r}
ggplot(cosmatdf) +
  geom_point(aes(x=reorder(username, -corr_may), y= corr_may)) + 
  coord_flip() +
  xlab("MP username") +
  ylab("Cosine similarity score") + 
  theme_minimal()

```

Combining these steps into a single `for` loop, we can see how our different similarity measures of interest compare. 

```{r}

#specify different similarity measures to explore
methods <- c("correlation", "cosine", "dice", "edice")

#create empty dataframe
testdf_all <- data.frame()

#gen for loop across methods types
for (i in seq_along(methods)) {
  
  #pass method to character string object
  sim_method <- methods[[i]]
  
  #estimate similarity, grouping by username
  test <- dfmat %>%
    dfm_group(groups = username) %>%
    textstat_simil(margin = "documents", method = sim_method) #specify method here as character object created above
  
  testm <- as.matrix(test) #convert to a matrix
  
  #generate data frame keeping only the row for Theresa May
  testdf <- as.data.frame(testm[23, c(1:22, 24)])
  
  #rename column
  colnames(testdf) <- "corr_may"
  
  #create column variable from rownames
  testdf <- tibble::rownames_to_column(testdf, "username")
  
  #record method in new column variable
  testdf$method <- sim_method

  #bind all together
  testdf_all <- rbind(testdf_all, testdf)  
  
}

#create variable (for viz only) that is mean of similarity scores for each MP
testdf_all <- testdf_all %>%
  group_by(username) %>%
  mutate(mean_sim = mean(corr_may))

ggplot(testdf_all) +
  geom_point( aes(x=reorder(username, -mean_sim), y= corr_may, color = method)) + 
  coord_flip() +
  xlab("MP username") +
  ylab("Similarity score") + 
  theme_minimal()

```

## Complexity

We now move to document-level measures of text characteristics. And here we will focus on the paper by @schoonvelde_liberals_2019. 

We will be using a subset of these data, taken from EU speeches given by four politicians. These are provided by the authors at [https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/S4IZ8K](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/S4IZ8K).

We can load the data as follows.

```{r}
speeches  <- readRDS(gzcon(url("https://github.com/cjbarrie/CTA-ED/blob/main/data/comparison-complexity/speeches.rds?raw=true")))
```

And we can take a look at what the data contains below.

```{r}
head(speeches)
```

The data contain speeches by four different politicians, each of whom are positioned at different points on a liberal-conservative scale.

We can then calculate the Flesch-Kincaid readability/complexity score with the `quanteda.textstats` package like so.

```{r}
speeches$flesch.kincaid <- textstat_readability(speeches$text, measure = "Flesch.Kincaid")

# returned as quanteda data.frame with document-level information;
# need just the score:
speeches$flesch.kincaid <- speeches$flesch.kincaid$Flesch.Kincaid
```

We want this information aggregated over each of our politicians: Gordon Brown, Jose Zapatero", David Cameron, and Mariano Rajoy. These are recorded in the data under a column called "speaker."

```{r}
#get mean and standard deviation of Flesch-Kincaid, and N of speeches for each speaker
sum_corpus <- speeches %>%
  group_by(speaker) %>%
  summarise(mean = mean(flesch.kincaid, na.rm=TRUE),
                   SD=sd(flesch.kincaid, na.rm=TRUE),
                   N=length(speaker))

# calculate standard errors and confidence intervals
sum_corpus$se <- sum_corpus$SD / sqrt(sum_corpus$N)
sum_corpus$min <- sum_corpus$mean - 1.96*sum_corpus$se
sum_corpus$max <- sum_corpus$mean + 1.96*sum_corpus$se
```

And this gives us data in tidy format that looks like so. 

```{r}
sum_corpus
```

Which we can then plot---and we see that our results look like those in Figure 1 of the published article by @schoonvelde_liberals_2019. 


```{r}

ggplot(sum_corpus, aes(x=speaker, y=mean)) +
  geom_bar(stat="identity") + 
  geom_errorbar(ymin=sum_corpus$min,ymax=sum_corpus$max, width=.2) +
  coord_flip() +
  xlab("") +
  ylab("Mean Complexity") + 
  theme_minimal() + 
  ylim(c(0,20))

```

## Exercises

### question 1: distance measures
1. Compute distance measures such as "euclidean" or "manhattan" for the MP tweets as above, comparing between tweets by MPs and tweets by PM, Theresa May. 

First we estimate distance measures, grouping by username. We compute two matrices of distance using  the euclidian and the manhattan measures respectively.

```{r}
# with euclidian method
euc_dist <- dfmat %>%
  dfm_group(groups = username) %>%
  textstat_dist(margin = "documents", method = "euclidean") %>% # use textstat_dist(), specify method 
  as.matrix() # turn into matrix output

# with 'manhattan' method
man_dist <- dfmat %>%
  dfm_group(groups = username) %>%
  textstat_dist(margin = "documents", method = "manhattan") %>% # use textstat_dist(), specify method
  as.matrix() # matrix output
```

We want to compare MPs tweets to the Prime Minister's tweets (then Theresa May), so we trim the data accordingly
```{r}
# generate data frames keeping only the row for Theresa May
euc_dist_may <- as.data.frame(euc_dist[23, c(1:22, 24)])
man_dist_may <- as.data.frame(man_dist[23, c(1:22, 24)])
```

We then rename the distance column with an appropriate name and convert row names to a column variable so that we have cells containing information on the MP to which the distance measure refers. 

```{r}
#rename distance column
colnames(euc_dist_may) <- "dist_may"
colnames(man_dist_may) <- "dist_may"
  
#create column variable from rownames
euc_dist_may <- tibble::rownames_to_column(euc_dist_may, "username")
man_dist_may <- tibble::rownames_to_column(man_dist_may, "username")

# create new column variable to identify the distance measure used
euc_dist_may <- euc_dist_may %>%
  mutate(dist_method = "euclidean")
man_dist_may <- man_dist_may %>%
  mutate(dist_method = "manhattan")
```

Now we have our data in tidy format. Each dataset has the same column names and thanks to the dist_method column we can easily identify which measures was computed through which method. We can now bind the two dataframes so we have both distance measures in a single dataframes.

```{r}
# bind the two dataframes
dist_may <- euc_dist_may %>%
  rbind(man_dist_may)

# check to see it looks right
head(dist_may)
table(dist_may$dist_method)
# all good
```

We can then plot like so.
```{r}
# plot
ggplot(dist_may) +
  geom_point(aes(x=reorder(username, -dist_may), y= dist_may, colour=dist_method)) + 
  coord_flip() +
  xlab("MP username") +
  ylab("Distance") + 
  labs(colour = "Distance measure") +
  theme_minimal()
```
Penny Mordant's and Lewis Brandon's tweets seem to be furthest away from Theresa May's tweets. This is consistent across both euclidean and manhattan measures.

### question 2: complexity measures
2. Estimate at least three other complexity measures for the EU speeches as above. Consider how the results compare to the Flesch-Kincaid measure used in the article by @schoonvelde_liberals_2019.

I chose mean sentence length, mean word syllables, and the Flesch score.
```{r}
# specify different complexity measures to explore
speeches <- speeches %>%
  mutate(meansentencelength = textstat_readability(text, measure = "meanSentenceLength"),
         meanwordsyllables = textstat_readability(text, measure = "meanWordSyllables"),
         flesch = textstat_readability(text, measure = "Flesch"))

# keep only the scores for the complexity measures (not the document identifier) 
speeches <- speeches %>%
  mutate(meansentencelength = meansentencelength$meanSentenceLength,
         meanwordsyllables = meanwordsyllables$meanWordSyllables,
         flesch = flesch$Flesch)
```

We aggregate the information over each of our politicians: Gordon Brown, Jose Zapatero", David Cameron, and Mariano Rajoy. These are recorded in the data under a column called "speaker."

```{r}
#get mean and standard deviation for each, and N of speeches for each speaker and for each measure
# to do this efficiently I use "across()" syntax which allows us to run the same command over multiple columns.

sum_corpus <- speeches %>%
  group_by(speaker) %>%
  summarise(
    across(c(flesch.kincaid, meansentencelength, meanwordsyllables, flesch), # across these 4 columns
           ~mean(.x, na.rm=TRUE), # run this function
           .names = "mean_{.col}"), # put results into new variables called "mean_[initial column name]
    across(c(flesch.kincaid, meansentencelength, meanwordsyllables, flesch), 
           ~sd(.x, na.rm=TRUE), .names = "sd_{.col}"),
    n=length(speaker))


# calculate standard errors and confidence intervals for each speaker and measure
# here I use the simpler method of duplicating code chunks

sum_corpus <- sum_corpus %>%
  mutate(# standard errors
         se_flesch.kincaid = sd_flesch.kincaid/sqrt(n),
         se_meansentencelength = sd_meansentencelength/sqrt(n),
         se_meanwordsyllables = sd_meanwordsyllables/sqrt(n),
         se_flesch = sd_flesch/sqrt(n),
         # lower brackets of confidence intervals
         min_flesch.kincaid = mean_flesch.kincaid - 1.96*se_flesch.kincaid,
         min_meansentencelength = mean_meansentencelength - 1.96*se_meansentencelength,
         min_meanwordsyllables = mean_meanwordsyllables - 1.96*se_meanwordsyllables,
         min_flesch = mean_flesch - 1.96*se_flesch,
         # upper brackets of confidence intervals
         max_flesch.kincaid = mean_flesch.kincaid + 1.96*se_flesch.kincaid,
         max_meansentencelength = mean_meansentencelength + 1.96*se_meansentencelength,
         max_meanwordsyllables = mean_meanwordsyllables + 1.96*se_meanwordsyllables,
         max_flesch = mean_flesch + 1.96*se_flesch)
```

Our data now looks like this:
```{r}
head(sum_corpus)
```

In order to plot, we need to do just a bit of data wrangling: namely we need to pivot our data into long format. The pivot_longer() command from the tidyverse package does just this.
```{r}
# pivot means
sum_corpus_long_mean <- sum_corpus %>%
  select(speaker, n, matches("mean_")) %>%
  pivot_longer(cols = !c(speaker, n),
               names_to = c("method"),
               names_pattern = "mean_(.*)",
               values_to = "mean")

# pivot lower confidence interval brackets
sum_corpus_long_min <- sum_corpus %>%
  select(speaker, n, matches("min_")) %>%
  pivot_longer(cols = !c(speaker, n),
               names_to = c("method"),
               names_pattern = "min_(.*)",
               values_to = "min")

# pivot upper confidence interval brackets
sum_corpus_long_max <- sum_corpus %>%
  select(speaker, n, matches("max_")) %>%
  pivot_longer(cols = !c(speaker, n),
               names_to = c("method"),
               names_pattern = "max_(.*)",
               values_to = "max")

# merge all
sum_corpus_long <- sum_corpus_long_mean %>%
  full_join(sum_corpus_long_min, by=c("speaker", "n", "method")) %>%
  full_join(sum_corpus_long_max, by=c("speaker", "n", "method"))
```


We can then plot results like so
```{r}
sum_corpus_long %>%
  ggplot(aes(x=speaker, y=mean, fill=speaker)) +
  geom_bar(stat = "identity") + 
  geom_errorbar(ymin=sum_corpus_long$min,ymax=sum_corpus_long$max, width=.2) +
  facet_wrap(~method) +
  coord_flip() +
  xlab("") +
  ylab("Mean (readability measures)") + 
  guides(fill=FALSE) +
  theme_minimal() 
```
We get quite different results across different measures. Where the Flesch-kincaid gave low readability to D. Cameron, the Flesch measure scores his speeches as more readable than the other 3. The Flesch-kincaid scores seem overall consistent with mean sentence length, whereas the Flesch scores arent. This could be a problem as we would be reaching different conclusions with different measures. 



3. (Advanced---optional) Estimate similarity scores between the MP tweets and the PM tweets for each week contained in the data. Plot the results. 